---
title: "Housing Starts Analysis — Pre-2006 Regime"
author: "Maximilian Van Zandt"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(astsa)
library(readr)
library(forecast)
library(tseries)
```

```{r load-data}
data <- read_csv("HOUST.csv")
data_ts <- ts(data$HOUST, start = c(1959, 1), frequency = 12)
```

#### Pre-2006

I switched up so just dealing with non-seasonal for right now, then checking ACF for spikes. I think breaking it down like this should help instead of trying to do all at once with an assumed seasonal pattern. I don't want to assume 60 or 12-month seasonality and just blindly generate plots. So,

1st: we decide if differencing/log transform is appropriate

2nd: We look at seasonal components

```{r pre2006-transform}
#Pre-Jan 2006
# Train/test split
train <- window(data_ts, end = c(2006, 1))
test  <- window(data_ts, start = c(2006, 2))

#Original (nonlinear trend, non-stationary)
housing_pre2006 <- train
plot(housing_pre2006, main="Housing Starts: Pre-2006", ylab="Housing Starts")
#Differencing
diff_pre2006 <- diff(housing_pre2006)
plot(diff_pre2006, main="Differenced Housing: Pre-2006")
#Log original
log_pre2006 <- log(housing_pre2006)
plot(log_pre2006)
#Difference the log
diff_log_pre2006 <- diff(log_pre2006)
plot(diff_log_pre2006, main="Differenced Log Housing: Pre-2006")
```

Difference of log is much much better than difference of original data. Below, I'm testing whether or not we have that 60-month cycle or that 12-month cycle by plotting the acf all the way up to lag 120. I thought 60 because of the periodogram we saw earlier but the spikes are clearly at 12 and the ACF values indicate 12.

```{r pre2006-acf}
acf2(diff(log(housing_pre2006)), max.lag=120)
```

Now, I'm checking seasonal differencing of the differenced log data just to see if that makes a difference.

```{r pre2006-seasonal-diff}
#Recall diff_pre2006 <- diff(log_pre2006)
#hardcoding it bc the tutorial hardcodes
both_diff_pre2006 <- diff(diff(log(housing_pre2006), 12))
plot(both_diff_pre2006, main="First + Seasonal Differenced: Pre-2006")
acf2(both_diff_pre2006, max.lag=120)
#Stationarity check
adf.test(diff_log_pre2006)
adf.test(both_diff_pre2006)
```

Results are inconclusive, but lean toward skipping the seasonal differencing in favor of simply differencing the log. The models below will include either seasonal and non-seasonal differencing and both and I'll compare from there.

```{r pre2006-sarima-candidates}
fit1 <- sarima(log_pre2006, 0, 0, 1, 0, 0, 1, 12)
fit2 <- sarima(log_pre2006, 0, 0, 1, 1, 0, 1, 12)
fit3 <- sarima(log_pre2006, 1, 0, 0, 1, 0, 1, 12)
fit4 <- sarima(log_pre2006, 1, 0, 1, 1, 0, 1, 12)
fit_pre2006 <- auto.arima(log(housing_pre2006), seasonal=TRUE, stepwise=FALSE,
                           approximation=FALSE, max.p=1, max.q=1, max.P=1, max.Q=1)
summary(fit_pre2006)
```

Model 4 wins!

```{r pre2006-forecast}
fitx <- Arima(log_pre2006, order = c(1,0,1), seasonal = c(1,0,1))

h <- length(test)
fc <- forecast(fitx, h = h)

# Back-transform to original scale
pred_level  <- exp(fc$mean)
lower_level <- exp(fc$lower[,2])  # 95% lower
upper_level <- exp(fc$upper[,2])  # 95% upper

# Plot: actual vs predicted (with 95% band)
plot(data_ts, main = "Housing Starts: Actual vs Forecast (level scale) (1,0,1)x(1,0,1)[12]",
     ylab = "Housing Starts", xlab = "Time", col = "black", lwd = 1,
     ylim=c(0,4000))
lines(pred_level, col = "red", lwd = 1)
lines(lower_level, col = "grey", lty = 2)
lines(upper_level, col = "grey", lty = 2)
legend("topleft",
       legend = c("Actual", "Forecast", "95% PI"),
       col = c("black", "red", "grey"),
       lty = c(1, 1, 2), lwd = c(2, 2, 1), bty = "n")
```

Reproducing the periodograms to see what we have now.

```{r pre2006-periodogram}
# Raw and smoothed periodograms
log_pre2006.per <- mvspec(log_pre2006)
k <- kernel("modified.daniell", c(3, 3))
log_pre2006.smooth <- mvspec(log_pre2006, k)

#Dominant frequency
n <- length(log_pre2006)
np <- nextn(n)
max_index <- which.max(log_pre2006.smooth$spec)
omega_star <- (max_index - 1) / np
period_star <- 1 / omega_star

cat("Dominant period:", round(period_star, 1), "months (",
    round(period_star/12, 1), "years)\n")

# Annual cycle power
annual_freq <- 1/12
annual_index <- which.min(abs(log_pre2006.smooth$freq - annual_freq))
annual_power <- log_pre2006.smooth$spec[annual_index]

cat("Annual cycle power ratio:", round(annual_power / log_pre2006.smooth$spec[max_index], 2), "\n")
```

```{r pre2006-spectral-setup}
# Data setup
train <- log_pre2006
n_train <- length(train)
t_train <- seq_len(n_train)

## 1. Detrend + demean training data
trend_fit <- lm(as.numeric(train) ~ t_train)
r_train <- resid(trend_fit)
r_train <- r_train - mean(r_train)

## 2. Smoothed spectral density (training only)
k <- kernel("modified.daniell", c(2, 2))
spec_obj <- mvspec(r_train, kernel = k, plot = TRUE)
spec_values <- spec_obj$spec
freq_values <- spec_obj$freq
df <- spec_obj$df

## 3. Dominant frequency
idx <- which.max(spec_values)
omega_star <- freq_values[idx]
period_star <- 1 / omega_star
f_hat <- spec_values[idx]

cat("Dominant period:", round(period_star, 1), "months (",
    round(period_star/12, 1), "years)\n")
cat("Dominant spectral power:", round(f_hat, 6), "\n")

## 4. Annual cycle comparison
annual_freq <- 1/12
annual_index <- which.min(abs(freq_values - annual_freq))
annual_power <- spec_values[annual_index]

cat("\nAnnual frequency:", freq_values[annual_index], "\n")
cat("Annual spectral power:", round(annual_power, 6), "\n")
cat("Annual/Dominant ratio:", round(annual_power / f_hat, 2), "\n")
```

```{r pre2006-spectral-forecast}
# Define test period (2006-2024)
test <- window(data_ts, start=c(2006,1))
n_test <- length(test)
t_test <- (n_train + 1):(n_train + n_test)

## 4. Fit sinusoid at dominant frequency
X_train <- cbind(
  cos(2 * pi * omega_star * t_train),
  sin(2 * pi * omega_star * t_train)
)
fit_fourier <- lm(r_train ~ X_train)

## 5. Forecast sinusoid into test period
X_test <- cbind(
  cos(2 * pi * omega_star * t_test),
  sin(2 * pi * omega_star * t_test)
)
r_hat_test <- cbind(1, X_test) %*% coef(fit_fourier)

## Add back trend
trend_test <- predict(
  trend_fit,
  newdata = data.frame(t_train = t_test)
)
log_pred <- as.numeric(r_hat_test) + trend_test

## 6. Spectral-based variance estimate
sigma2_hat <- f_hat
alpha <- 0.05
z <- qnorm(1 - alpha / 2)
se_pred <- sqrt(sigma2_hat)
lower <- log_pred - z * se_pred
upper <- log_pred + z * se_pred

## 7. Collect results
forecast_df <- ts(
  cbind(
    actual = log(test),
    forecast = log_pred,
    lower = lower,
    upper = upper
  ),
  start = start(test),
  frequency = frequency(test)
)
forecast_df
```

```{r pre2006-fitted-values}
## In-sample fitted values (log scale)
r_hat_train <- cbind(1, X_train) %*% coef(fit_fourier)
trend_train <- fitted(trend_fit)
log_fitted <- as.numeric(r_hat_train) + trend_train
fitted_level <- exp(log_fitted)
fitted_ts <- ts(
  fitted_level,
  start = start(train),
  frequency = frequency(train)
)
```

```{r pre2006-spectral-plot}
pred_level  <- exp(log_pred)
lower_level <- exp(lower)
upper_level <- exp(upper)
pred_ts  <- ts(pred_level,  start = start(test), frequency = frequency(test))
lower_ts <- ts(lower_level, start = start(test), frequency = frequency(test))
upper_ts <- ts(upper_level, start = start(test), frequency = frequency(test))

smoothed_data_ts <- ma(data_ts, order = period_star)

plot(
  smoothed_data_ts,
  main = "Housing Starts: Spectral Fit and Forecast (level scale)",
  ylab = "Housing Starts",
  xlab = "Time",
  col = "black",
  lwd = 1,
  ylim = c(0, 4000)
)

## In-sample fitted (training)
lines(fitted_ts, col = "blue", lwd = 1)

## Out-of-sample forecast
lines(pred_ts,  col = "red",  lwd = 1)

legend(
  "topleft",
  legend = c("Actual (smoothed)", "Fitted (Train)", "Spectral Forecast"),
  col = c("black", "blue", "red"),
  lty = c(1, 1, 1, 2),
  lwd = c(2, 2, 2, 1),
  bty = "n"
)
```
